Title Goes Here

### 2025-04-21 analysis/paper/paper.qmd
Plots in main paper improved. Removed Cu from the PCA analysis

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(ggpmisc)
library(magick)

@@

# Read individual xrd files
sp15_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP15_xrd.csv')
sp58_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP58_xrd.csv')
sp62_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP62_xrd.csv')
sp69_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP69_xrd.csv')

@@

sp15_plot <- sp15_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000) |> 
  select(-Quartz_nb, -Dolomite_nb) |>
  pivot_longer(cols = c(Difference, Background, Observed, Quartz, Dolomite), 
               names_to = "Phase", 
               values_to = "Count") |> 
  mutate(Phase = factor(Phase, levels = c("Observed", "Quartz", "Dolomite", "Background", "Difference"))) |>
  ggplot(aes(x = Angle, y = Count, colour = Phase)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity") +
  theme_minimal() +
  coord_cartesian(ylim = c(-5000, 55000)) +
  geom_point(data = sp15_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  geom_label(data = sp15_marks, aes(x = Angle, y = Intensity, label = d), 
            size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  scale_colour_manual(values = c("lightsteelblue3", "salmon", "steelblue1" ,"orchid","snow4")) +
  annotate(geom = "table", x = 75, y = 55000, label = list(cbind(sp15_perc)))

@@

sp58_plot <- sp58_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000) |> 
  select(-Quartz_nb, -Calcite_nb) |>
  pivot_longer(cols = c(Difference, Background, Observed, Quartz, Calcite), 
               names_to = "Phase", 
               values_to = "Count") |> 
  mutate(Phase = factor(Phase, levels = c("Observed", "Quartz", "Calcite", "Background", "Difference"))) |>
  ggplot(aes(x = Angle, y = Count, colour = Phase)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity") +
  theme_minimal() +
  coord_cartesian(ylim = c(-5000, 50000)) +
  geom_point(data = sp58_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  geom_label(data = sp58_marks, aes(x = Angle, y = Intensity, label = d), 
            size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  scale_colour_manual(values = c("lightsteelblue3", "salmon", "darkseagreen" ,"orchid","snow4")) +
  annotate(geom = "table", x = 75, y = 50000, label = list(cbind(sp58_perc)))

# SP62 Oliveiras
sp62_marks <- xrd_marks |> 
  filter(Sample == "SP62")
sp62_perc <- xrd_perc |> 
  filter(Sample == "SP62") |> 
  select(-Sample, -ESD)

sp62_plot <- sp62_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000) |> 
  select(-Quartz_nb) |> 
  pivot_longer(cols = c(Difference, Background, Observed, Quartz), 
               names_to = "Phase", 
               values_to = "Count") |> 
  mutate(Phase = factor(Phase, levels = c("Observed", "Quartz", "Background", "Difference"))) |>
  ggplot(aes(x = Angle, y = Count, colour = Phase)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity") +
  theme_minimal() +
  coord_cartesian(ylim = c(-5000, 50000)) +
  geom_point(data = sp62_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  geom_label(data = sp62_marks, aes(x = Angle, y = Intensity, label = d), 
            size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  scale_colour_manual(values = c("lightsteelblue3", "salmon", "orchid","snow4")) +
  annotate(geom = "table", x = 75, y = 50000, label = list(cbind(sp62_perc)))

@@

sp69_plot <- sp69_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000) |> 
  select(-Quartz_nb, -Calcite_nb) |> 
  pivot_longer(cols = c(Difference, Background, Observed, Quartz, Calcite), 
               names_to = "Phase", 
               values_to = "Count") |> 
  mutate(Phase = factor(Phase, levels = c("Observed", "Quartz", "Calcite", "Background", "Difference"))) |>
  ggplot(aes(x = Angle, y = Count, colour = Phase)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity") +
  theme_minimal() +
  coord_cartesian(ylim = c(-5000, 50000)) +
  geom_point(data = sp69_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  geom_label(data = sp69_marks, aes(x = Angle, y = Intensity, label = d), 
            size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  scale_colour_manual(values = c("lightsteelblue3", "salmon", "darkseagreen" ,"orchid","snow4")) +
  annotate(geom = "table", x = 75, y = 50000, label = list(cbind(sp69_perc)))

pxrf_raw <- read_csv('data/pxrf_data_dec.csv')

@@

@@

```{r}
#| label: pxrf-fig-lod
#| fig-cap: A barplot.

@@

# Reshape the data into a wide format using pivot_wider
lod_tab_table <- lod_tab_summary |>
  pivot_wider(names_from = Value, values_from = Count, values_fill = list(Count = 0)) |> 
  filter(Success > 80) |> 
  select(Element, Success, "< LOD")
lod_tab_table_long <- lod_tab_table |> 
  pivot_longer(cols = c(Success, `< LOD`), names_to = "Metric", values_to = "Count") |> 
  mutate(Metric = factor(Metric, levels = c("Success", "< LOD"))) 
lod_tab_table_long |> 
  ggplot(aes(x = Element, y = Count, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Success and < LOD", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Success" = "#99D3A7", "< LOD" = "#ED7673"))

# Create the bar plot
ggplot(lod_tab_summary, aes(x = Element, y = Count, fill = Value)) +
  geom_bar(stat = "identity", position = "stack") +  # Use stat = "identity" to use summarized counts
  scale_fill_manual(values = c("Success" = "darkseagreen", "Success 0" = "lightgoldenrod1", "< LOD" = "indianred3")) +
  labs(title = "Summarized Success and LOD Counts by Element",
       y = "Count", 
       fill = " ") +
  theme_minimal() +
 # facet_wrap(~ Element, scales = "free_y") +  # Facet by Element with independent y-axes
  theme(axis.text.x = element_text(hjust = 1),
        axis.title.x = element_blank()) +
  #geom_text(data = lod_tab_summary_filtered, aes(label = Count), position=position_stack(vjust=0.5), colour="white", size = 3) +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, max(lod_tab_summary$Count), by = 20))

@@

## a) PCA geological regional cherts
pca_db_algarve <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, S, Cl, K2O, CaO, TiO2, Fe2O3) |>
  filter(`Formation/Type` %in% c("LJW", "MJE", "UJE", "UJW", "TW"))
# Convert "< LOD" values to NA
pca_db_algarve[pca_db_algarve == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_algarve$`Formation/Type` <- as.factor(pca_db_algarve$`Formation/Type`)  # Ensure it's a factor
group_var_algarve <- pca_db_algarve$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_alg <- pca_db_algarve %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_alg <- complete.cases(df_numeric_alg)  # Identify complete rows
df_numeric_alg <- df_numeric_alg[complete_cases_alg, ]  # Remove NA rows
group_var_algarve <- group_var_algarve[complete_cases_alg]  # Filter group_var accordingly
# Scale values
df_scaled_alg <- scale(df_numeric_alg)
# Ensure row names are properly set
rownames(df_scaled_alg) <- make.names(group_var_algarve, unique = TRUE)
# Perform PCA
pca_result_alg <- PCA(df_scaled_alg, graph = FALSE)
# Summary of PCA
summary(pca_result_alg)
fviz_eig(pca_result_alg, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_alg <- fviz_pca_biplot(pca_result_alg, 
                label = "var", 
                habillage = group_var_algarve,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.95) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_alg <- data.frame(pca_result_alg$ind$coord, Group = group_var_algarve)
# PCA plot with groups
scatter_alg <- ggplot(df_pca_alg, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

## b) PCA of geological cherts: local Lower Jurassic, Central Portugal and Cadiz
pca_db_nl <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, K2O, TiO2, Fe2O3, Cu, S) |> # Removing CaO which is high NA in CPT
  filter(`Formation/Type` %in% c("LJW", "CPT", "UCA"))
# Convert "< LOD" values to NA
pca_db_nl[pca_db_nl == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_nl$`Formation/Type` <- as.factor(pca_db_nl$`Formation/Type`)  # Ensure it's a factor
group_var_nl <- pca_db_nl$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_nl <- pca_db_nl %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_nl <- complete.cases(df_numeric_nl)  # Identify complete rows
df_numeric_nl <- df_numeric_nl[complete_cases_nl, ]  # Remove NA rows
group_var_nl <- group_var_nl[complete_cases_nl]  # Filter group_var accordingly
# Scale values
df_scaled_nl <- scale(df_numeric_nl)
# Ensure row names are properly set
rownames(df_scaled_nl) <- make.names(group_var_nl, unique = TRUE)
# Perform PCA
pca_result_nl <- PCA(df_scaled_nl, graph = FALSE)
# Summary of PCA
summary(pca_result_nl)
fviz_eig(pca_result_nl, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_nl <- fviz_pca_biplot(pca_result_nl, 
                label = "var", 
                habillage = group_var_nl,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.95) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_nl <- data.frame(pca_result_nl$ind$coord, Group = group_var_nl)
# PCA plot with groups
scatter_nl <- ggplot(df_pca_nl, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

## c) PCA archaeological T6 compared to local Lower Jurassic and Central Portugal
pca_db_comp <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, K2O, TiO2, Fe2O3, Cu, S) |> # Removing CaO which is high NA in CPT
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "TYPE 1" ~ "T1",
    `Formation/Type` == "TYPE 2" | `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "T2",
    `Formation/Type` == "TYPE 3" ~ "T3",
    `Formation/Type` == "TYPE 4" ~ "T4",
    `Formation/Type` == "TYPE 5" ~ "T5",
    `Formation/Type` == "TYPE 6" | `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "T6",
    `Formation/Type` == "TYPE 7" ~ "T7",
    `Formation/Type` == "TYPE 8" ~ "T8",
    `Formation/Type` == "TYPE 9" ~ "T9",
    `Formation/Type` == "TYPE 10" ~ "T10",
    `Formation/Type` == "TYPE 11" ~ "T11",
    TRUE ~ `Formation/Type`
  )) |> 
  filter(`Formation/Type` %in% c("LJW","CPT", "T6"))
# Convert "< LOD" values to NA
pca_db_comp[pca_db_comp == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_comp$`Formation/Type` <- as.factor(pca_db_comp$`Formation/Type`)  # Ensure it's a factor
group_var_comp <- pca_db_comp$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_comp <- pca_db_comp %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_comp <- complete.cases(df_numeric_comp)  # Identify complete rows
df_numeric_comp <- df_numeric_comp[complete_cases_comp, ]  # Remove NA rows
group_var_comp <- group_var_comp[complete_cases_comp]  # Filter group_var accordingly
# Scale values
df_scaled_comp <- scale(df_numeric_comp)
# Ensure row names are properly set
rownames(df_scaled_comp) <- make.names(group_var_comp, unique = TRUE)
# Perform PCA
pca_result_comp <- PCA(df_scaled_comp, graph = FALSE)
# Summary of PCA
summary(pca_result_comp)
fviz_eig(pca_result_comp, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_comp <- fviz_pca_biplot(pca_result_comp, 
                label = "var", 
                habillage = group_var_comp,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.80) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_comp <- data.frame(pca_result_comp$ind$coord, Group = group_var_comp)
# PCA plot with groups
scatter_comp <- ggplot(df_pca_comp, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

>>>>

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(ggpmisc)
library(magick)
library(ggpubr)

@@

# Read individual xrd files
sp15_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP15_xrd.csv')
sp58_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP58_xrd.csv')
sp69_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP69_xrd.csv')

@@

sp15_plot <- sp15_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000) |> 
  select(-Quartz_nb, -Dolomite_nb) |>
  pivot_longer(cols = c(Difference, Background, Observed, Quartz, Dolomite), 
               names_to = "Phase", 
               values_to = "Count") |> 
  mutate(Phase = factor(Phase, levels = c("Observed", "Quartz", "Dolomite", "Background", "Difference"))) |>
  ggplot(aes(x = Angle, y = Count, colour = Phase)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity (counts)") +
  theme_minimal() +
  coord_cartesian(ylim = c(-5000, 55000)) +
  geom_point(data = sp15_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  geom_label(data = sp15_marks, aes(x = Angle, y = Intensity, label = d), 
            size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  scale_colour_manual(values = c("lightsteelblue3", "salmon", "steelblue1" ,"orchid","snow4")) +
  annotate(geom = "table", x = 75, y = 55000, label = list(cbind(sp15_perc)))

@@

sp58_plot <- sp58_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000) |> 
  select(-Quartz_nb, -Calcite_nb) |>
  pivot_longer(cols = c(Difference, Background, Observed, Quartz, Calcite), 
               names_to = "Phase", 
               values_to = "Count") |> 
  mutate(Phase = factor(Phase, levels = c("Observed", "Quartz", "Calcite", "Background", "Difference"))) |>
  ggplot(aes(x = Angle, y = Count, colour = Phase)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity (counts)") +
  theme_minimal() +
  coord_cartesian(ylim = c(-5000, 50000)) +
  geom_point(data = sp58_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  geom_label(data = sp58_marks, aes(x = Angle, y = Intensity, label = d), 
            size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  scale_colour_manual(values = c("lightsteelblue3", "salmon", "darkseagreen" ,"orchid","snow4")) +
  annotate(geom = "table", x = 75, y = 50000, label = list(cbind(sp58_perc)))

@@

sp69_plot <- sp69_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000) |> 
  select(-Quartz_nb, -Calcite_nb) |> 
  pivot_longer(cols = c(Difference, Background, Observed, Quartz, Calcite), 
               names_to = "Phase", 
               values_to = "Count") |> 
  mutate(Phase = factor(Phase, levels = c("Observed", "Quartz", "Calcite", "Background", "Difference"))) |>
  ggplot(aes(x = Angle, y = Count, colour = Phase)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity (counts)") +
  theme_minimal() +
  coord_cartesian(ylim = c(-5000, 50000)) +
  geom_point(data = sp69_marks, aes(x = Angle, y = Intensity + 600, colour = Phase), size = 3, shape = 18) +
  geom_label(data = sp69_marks, aes(x = Angle, y = Intensity, label = d), 
            size = 3, fill = "white", label.size = 0.2, vjust = -1, show.legend = FALSE) +
  scale_colour_manual(values = c("lightsteelblue3", "salmon", "darkseagreen" ,"orchid","snow4")) +
  annotate(geom = "table", x = 75, y = 50000, label = list(cbind(sp69_perc)))

@@

pxrf_raw <- read_csv('analysis/data/raw_data/pxrf_data/pxrf_data_dec.csv')

@@

```{r}
#| label: pxrf-lod

@@

# Reshape the data into a wide format using pivot_wider
lod_tab_table <- lod_tab_summary |>
  pivot_wider(names_from = Value, values_from = Count, values_fill = list(Count = 0)) |> 
  filter(Success > 80) |> 
  select(Element, Success, "< LOD")
lod_tab_table_long <- lod_tab_table |> 
  pivot_longer(cols = c(Success, `< LOD`), names_to = "Metric", values_to = "Count") |> 
  mutate(Metric = factor(Metric, levels = c("Success", "< LOD"))) 
succ_barplot <- lod_tab_table_long |> 
  ggplot(aes(x = Element, y = Count, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Success and < LOD", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Success" = "#99D3A7", "< LOD" = "#ED7673"))

```{r}
#| label: boxplots

# Boxplot data preparation
boxplots_pxrf <- pxrf_analysis |> 
  select(-Sample_ID, -pXRF_ID) |> 
  filter(!`Formation/Type` %in% c("TW", "T1", "T9", "T10", "T11")) |> # Removing chalcedony
  mutate(across(-`Formation/Type`, ~ as.numeric(.)))
  #pivot_longer(cols = c("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
        # "Sr", "Zr"), names_to = "Variable", values_to = "Value")
boxplots_pxrf <- boxplots_pxrf |> 
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "T2" | `Formation/Type` == "T3" | `Formation/Type` == "T4" | `Formation/Type` == "T5" ~ "T2-5", TRUE ~ `Formation/Type`
  )) |> 
  mutate(`Formation/Type` = factor(`Formation/Type`, levels = c("LJW", "UJW", "MJE", "UJE", "CPT", "UCA", "T1", "T2-5", "T6", "T7", "T8")))

## SiO2
count_data_sio2 <- boxplots_pxrf |> # Count number of observations per group where Value > 0
  select(`Formation/Type`, SiO2) |> 
  group_by(`Formation/Type`) |> 
  filter(!is.na(SiO2)) |> 
  summarise(n = n(), .groups = "drop")

si_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = SiO2, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_sio2, aes(x = `Formation/Type`, y = 70, label = n), size = 3) +
  coord_cartesian(ylim = c(70, 100)) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

## CaO
count_data_cao <- boxplots_pxrf |> # Count number of observations per group where Value > 0
  select(`Formation/Type`, CaO) |> 
  group_by(`Formation/Type`) |> 
  filter(!is.na(CaO)) |> 
  summarise(n = n(), .groups = "drop")

ca_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = CaO, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_cao, aes(x = `Formation/Type`, y = -0.3, label = n), size = 3) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

## TiO2
count_data_tio2 <- boxplots_pxrf |> # Count number of observations per group where Value > 0
  select(`Formation/Type`, TiO2) |> 
  group_by(`Formation/Type`) |> 
  filter(!is.na(TiO2)) |> 
  summarise(n = n(), .groups = "drop")

ti_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = TiO2, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_tio2, aes(x = `Formation/Type`, y = -0.02, label = n), size = 3) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

## Fe2O3
count_data_fe2o3 <- boxplots_pxrf |> # Count number of observations per group where Value > 0
  select(`Formation/Type`, Fe2O3) |> 
  group_by(`Formation/Type`) |> 
  filter(!is.na(Fe2O3)) |> 
  summarise(n = n(), .groups = "drop")

fe_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = Fe2O3, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_fe2o3, aes(x = `Formation/Type`, y = 0, label = n), size = 3) +
  coord_cartesian(ylim = c(0, 0.75)) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

## a) PCA geological regional cherts
pca_db_algarve <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, S, Cl, K2O, CaO, TiO2, Fe2O3) |>
  filter(`Formation/Type` %in% c("LJW", "MJE", "UJE", "UJW", "TW"))
# Convert "< LOD" values to NA
pca_db_algarve[pca_db_algarve == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_algarve$`Formation/Type` <- as.factor(pca_db_algarve$`Formation/Type`)  # Ensure it's a factor
group_var_algarve <- pca_db_algarve$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_alg <- pca_db_algarve %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_alg <- complete.cases(df_numeric_alg)  # Identify complete rows
df_numeric_alg <- df_numeric_alg[complete_cases_alg, ]  # Remove NA rows
group_var_algarve <- group_var_algarve[complete_cases_alg]  # Filter group_var accordingly
# Scale values
df_scaled_alg <- scale(df_numeric_alg)
# Ensure row names are properly set
rownames(df_scaled_alg) <- make.names(group_var_algarve, unique = TRUE)
# Perform PCA
pca_result_alg <- PCA(df_scaled_alg, graph = FALSE)
# Summary of PCA
summary(pca_result_alg)
fviz_eig(pca_result_alg, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_alg <- fviz_pca_biplot(pca_result_alg, 
                label = "var", 
                habillage = group_var_algarve,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.95,
                palette = c("LJW" = "#EDA92B", "UJW" = "#FF8989", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "TW" = "#EC407A"),
                col.var = "#757575") +
  scale_shape_manual(values=c(16, 17, 8, 18, 15)) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_alg <- data.frame(pca_result_alg$ind$coord, Group = group_var_algarve)
# PCA plot with groups
scatter_alg <- ggplot(df_pca_alg, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

## b) PCA of geological cherts: local Lower Jurassic, Central Portugal and Cadiz
pca_db_nl <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, K2O, TiO2, Fe2O3, S) |> # Removing CaO which is high NA in CPT
  filter(`Formation/Type` %in% c("LJW", "CPT", "UCA"))
# Convert "< LOD" values to NA
pca_db_nl[pca_db_nl == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_nl$`Formation/Type` <- as.factor(pca_db_nl$`Formation/Type`)  # Ensure it's a factor
group_var_nl <- pca_db_nl$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_nl <- pca_db_nl %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_nl <- complete.cases(df_numeric_nl)  # Identify complete rows
df_numeric_nl <- df_numeric_nl[complete_cases_nl, ]  # Remove NA rows
group_var_nl <- group_var_nl[complete_cases_nl]  # Filter group_var accordingly
# Scale values
df_scaled_nl <- scale(df_numeric_nl)
# Ensure row names are properly set
rownames(df_scaled_nl) <- make.names(group_var_nl, unique = TRUE)
# Perform PCA
pca_result_nl <- PCA(df_scaled_nl, graph = FALSE)
# Summary of PCA
summary(pca_result_nl)
fviz_eig(pca_result_nl, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_nl <- fviz_pca_biplot(pca_result_nl, 
                label = "var", 
                habillage = group_var_nl,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.80,
                palette = c("LJW" = "#EDA92B", "CPT" = "#4EB3D3", "UCA" = "#93C54E"),
                col.var = "#757575") +
  scale_shape_manual(values=c(17, 16, 15)) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_nl <- data.frame(pca_result_nl$ind$coord, Group = group_var_nl)
# PCA plot with groups
scatter_nl <- ggplot(df_pca_nl, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")
